XXX in a Tracing JIT Compiler for Efficient Dynamic Languages

### 2011-03-25 talk/icooolps2011/paper.tex
improve intro and background

The PyPy project \cite{armin_rigo_pypys_2006} strives to be an environment where
complex dynamic languages can be efficiently implemented. The approach taken
when implement a language with PyPy is to write an interpreter for the language
in \emph{RPython}. RPython is a restricted subset of Python chosen in such a way
that it is possible to perform type inference on it. The interpreters in RPython
can therefore be translated to efficient C code.

A number of languages have been implemented with PyPy, most importantly a full
Python implementation, but also a Prolog interpreter \cite{XXX} and a Smalltalk
VM \cite{XXX}.

This translation to C code adds a number of implementation details into the
final executable that are not present in the interpreter implementation, such as
a garbage collector. The interpreter can therefore be kept free from low-level
implementation details. Another aspect of the final VM that is added
semi-automatically to the generated VM is a tracing JIT compiler.

@@

XXX citations
A recently popular approach to JIT compilers is that of tracing JITs. Tracing
JITs record traces of concrete execution paths through the program. Those
traces are therefore linear list of operations, which are optimized and then
get turned into machine code. To be able to do this recording, VMs with a
tracing JIT typically also contain an interpreter. After a user program is
started the interpreter is used until the most important paths through the user
program are turned into machine code.

@@

One disadvantage of tracing JITs which makes them not directly applicable to
PyPy is that they encode the language semantics. Since PyPy wants to be a
general framework, we want to reuse our tracer for different languages.
Therefore PyPy's JIT is a meta-tracer \cite{bolz_tracing_2009}. It does not
trace the execution of the user program, but instead traces the execution of
the \emph{interpreter} that is running the program. This means that the traces
it produces don't contain the bytecodes of the language in question, but
RPython-level operations that the interpreter did to execute the program.

On the other hand, the loops that are traced by the tracer are the loops in the
user program. This means that the tracer stops tracing after one iteration of
the loop in the user function that is being considered. At this point, it can
have traced many iterations of the interpreter main loop.

@@

XXX trace makes the object model operations explicit and transparent to the
optimizer

>>>>

The PyPy project \cite{armin_rigo_pypys_2006} strives to be an environment where
complex dynamic languages can be implemented efficiently. The approach taken
when implementing a language with PyPy is to write an interpreter for the language
in \emph{RPython}. RPython is a restricted subset of Python chosen in such a way
that it is possible to perform type inference on it. The interpreters in RPython
can therefore be translated to efficient C code.

A number of languages have been implemented with PyPy, most importantly a full
Python implementation, but also a Prolog interpreter
\cite{carl_friedrich_bolz_towards_2010} and a Smalltalk VM
\cite{carl_friedrich_bolz_back_2008}.

The translation of the interpreter to C code adds a number of implementation details into the
final executable that are not present in the interpreter implementation, such as
a garbage collector. The interpreter can therefore be kept free from low-level
implementation details. Another aspect of the final VM that is added
semi-automatically to the generated VM is a tracing JIT compiler.

@@

A recently popular approach to JIT compilers is that of tracing JITs. Tracing
JITs have their origin in the Dynamo project which used the for dynamic
assembler optimization \cite{XXX}. Later they were used for to implement
a lightweight JIT for Java \cite{XXX} and for dynamic languages such as
JavaScript \cite{XXX}.

A tracing JIT works by recording traces of concrete execution paths through the
program. Those
traces are therefore linear list of operations, which are optimized and then
get turned into machine code. To be able to do this recording, VMs with a
tracing JIT typically contain an interpreter. After a user program is
started the interpreter is used until the most important paths through the user
program are turned into machine code. The tracing JIT tries to produce traces
that correspond to loops in the traced program, but most tracing JITs now also
have support for tracing non-loops \cite{XXX}.

@@

One disadvantage of tracing JITs which makes them not directly applicable to
PyPy is that they need to encode the language semantics of the language they are
tracing. Since PyPy wants to be a
general framework, we want to reuse our tracer for different languages.
Therefore PyPy's JIT is a meta-tracer \cite{bolz_tracing_2009}. It does not
trace the execution of the user program, but instead traces the execution of
the \emph{interpreter} that is running the program. This means that the traces
it produces don't contain the bytecodes of the language in question, but
RPython-level operations that the interpreter did to execute the program.

Tracing through the execution of an interpreter has many advantages. It makes
the tracer, its optimizers and backends reusable for a variety of languages. The
language semantics do not need to be encoded into the JIT. Instead the tracer
just picks them up from the interpreter. XXX mention disadvantage of long
traces?

While the operations in a trace are those of the interpreter, the loops that are
traced by the tracer are the loops in the
user program. This means that the tracer stops tracing after one iteration of
the loop in the user function that is being considered. At this point, it can
have traced many iterations of the interpreter main loop.

@@
